A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2016-07-23 rigid_bodies.tex
Appendices almost concluded

%\documentclass[prb,preprint,preprintnumbers,amsmath,amssymb]{revtex4-1}
%\documentclass[prb,twocolumn,preprintnumbers,amsmath,amssymb]{revtex4-1}
\documentclass[aip,jcp,reprint,amsmath,amssymb,raggedbottom]{revtex4-1}

\usepackage[pdftex]{graphicx}
\usepackage{epstopdf}
\usepackage{caption}
\usepackage{subfig}
\usepackage[]{units}
\usepackage[]{natbib}
\usepackage[]{threeparttable}
\usepackage{booktabs}
\usepackage{multirow}

\newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}           % matrix symbol
\newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}           % vector symbol
\newcommand{\tr}[1]{#1^t}                               % transposition
\newcommand{\diff}[2]{\dfrac{\partial #1}{\partial #2}} % partial derivative

@@

In this section, we first review certain aspects of the kinematics of rigid bodies which are relevant to the present study. Unity quaternions are employed to represent the orientations of rigid bodies. Then, we introduce the use of a particular factorization of the rotation matrix, from which a mathematical relation arises between the rate of change of a four-dimensional quaternion and a three-dimensional angular velocity vector. An important feature of this relation is that it preserves the unit-norm constraint of the quaternion upon infinitesimal motion.

@@

Consider a rigid body composed of $n_p$ individual particles, with each particle $j$ having mass $M_j$ and being located at a Cartesian coordinate $\vt R_j$. The center of mass of this body is located at
\[
\vt r = \frac{1}{m}\sum_{j=1}^{n_p} M_j {\vt R}_j,
\]
where $m = \sum_{j=1}^{n_p} M_j$ is the total mass of the body. The position of each particle $j$ in a space-fixed frame of reference, whose origin coincides with the body's center of mass, is ${\vt \delta}_j = \tr{ [\begin{array}{ccc} {\delta_j}_x & {\delta_j}_y & {\delta_j}_z \end{array}]} = {\vt R}_j - {\vt r}$. In such frame, the symmetric inertia tensor for rotations of the body about the origin is\cite{Goldstein2002}
\[
{\mt I}^\ast = \sum_{j=1}^N M_j \left[ \begin{array}{ccc}
{\delta_j}_y^2 + {\delta_j}_z^2 & -{\delta_j}_x {\delta_j}_y & -{\delta_j}_x {\delta_j}_z \\
-{\delta_j}_y {\delta_j}_x & {\delta_j}_x^2 + {\delta_j}_z^2 & -{\delta_j}_y {\delta_j}_z \\
-{\delta_j}_z {\delta_j}_x & -{\delta_j}_z {\delta_j}_y & {\delta_j}_x^2 + {\delta_j}_y^2 \\
\end{array} \right].
\]

For every non-collinear rigid body, there exists a particular frame of reference in which its inertia tensor is diagonal. The corresponding axes are known as the principal axes and the formulation of the equations of motion in this frame of reference is notably simplified. Conversions from the space-fixed frame to such a body-fixed frame are performed by a rotation matrix $\mt A$ that satisfies the similarity relation\citep{Goldstein2002}
\[
{\mt I} = {\mt A} {\mt I}^\ast \tr{\mt A} = \left[ \begin{array}{ccc}
I_1 &   0 &   0 \\
  0 & I_2 &   0 \\
  0 &   0 & I_3 \\
\end{array} \right],
\]
where $I_1$, $I_2$, and $I_3$ (the principal moments of inertia) are the eigenvalues of ${\mt I}^\ast$ and the columns of $\tr{\mt A}$ (the principal axes of rotation) are their corresponding eigenvectors. Algorithms for calculating $\mt A$ from $\mt I^\ast$ can be found, for example, in Ref.~\onlinecite{Kopp2008}. Once this is done, the position of every particle $j$ in the body-fixed frame can be obtained by ${\vt d}_j = {\mt A} {\vt \delta}_j$. Since $\mt A$ represents an orthogonal transformation, a conversion in the opposite direction is done by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} This is so because $\mt A \tr{\mt A} = \mt 1_3$, where $\mt 1_k$ represents an identity matrix in $\mathbb{R}^{k \times k}$ . By differentiating both sides of this relation with respect to time, we conclude that ${\mt A} \tr{\dot{\mt A}} = -\dot{\mt A} \tr{\mt A}$, which means that $\mt A \tr{\dot{\mt A}}$ results in a skew-symmetric matrix. This product is related to $\vt \omega$, the angular velocity of the body in the body-fixed frame, by\cite{Haug1989}
\begin{equation}
\label{eq:relation_A_omega}
\mt S(\vt \omega) = \mt A \tr{\dot{\mt A}} = -\dot{\mt A} \tr{\mt A},
\end{equation}
where $\mt S(\cdot)$ is the skew-symmetric operator defined in Eq.~\ref{eq:operator_S}.

\subsection{The Rate of Change of Unit Quaternions}
\label{sec:eulerparameters}

Euler's rotation theorem establishes that an arbitrary rotation can be described using three independent parameters. One can use, for instance, the so-called Euler angles, but mathematical singularities arise when the equations of motion are described in such terms. Instead, we consider an alternative representation, in which the orientation of a rigid body is given by the Euler parameters $q_0$, $q_1$, $q_2$, and $q_3$, which form a quaternion $\vt q = \tr {[\begin{array}{cccc} q_0 & q_1 & q_2 & q_3 \end{array}]}$. The mathematical relation between the Euler angles and the Euler parameters requires that\cite{Goldstein2002}
\begin{equation}
\label{eq:norm_eq_1}
\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_0^2 + q_1^2 + q_2^2 + q_3^2 = 1.
\end{equation}

Therefore, $\vt q$ is restricted to the surface of a unit hypersphere in $\mathbb{R}^4$, which is itself a three-dimensional domain. Appendix~\ref{sec:quat_from_A} describes how one can compute the unit quaternion related to a given rotation matrix. Reversely, the matrix $\mt A$ can be directly computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
\begin{widetext}
\begin{equation}
\label{eq:A_from_q}
\mt A = \left[ \begin{array}{ccc}
q_0^2 + q_1^2 - q_2^2 - q_3^2 & 2(q_1 q_2 + q_0 q_3) & 2(q_1 q_3 - q_0 q_2) \\
2(q_1 q_2 - q_0 q_3) & q_0^2 - q_1^2 + q_2^2 - q_3^2 & 2(q_2 q_3 + q_0 q_1) \\
2(q_1 q_3 + q_0 q_2) & 2(q_2 q_3 - q_0 q_1) & q_0^2 - q_1^2 - q_2^2 + q_3^2  
\end{array} \right].
\end{equation}
\end{widetext}

The rate of change of a quaternion is denoted by $\dot{\vt q}$. Differentiating Eq.~\ref{eq:norm_eq_1} with respect to time yields
\begin{equation}
\label{eq:diff_qTq}
\tr{\vt q}\dot{\vt q} = 0,
\end{equation}
which implies that, for Eq.~\ref{eq:norm_eq_1} to be preserved upon an infinitesimal motion, $\dot{\vt q}$ must lie in the hyperplane orthogonal to $\vt q$.

Let us now introduce two new matrices that also depend on the Euler parameters, which are
\begin{subequations}
\label{eq:def_B_and_C}
\begin{align}
\mt B &= \left[
\begin{array}{rrrr}
-q_1 & -q_2 & -q_3 \\
 q_0 & -q_3 &  q_2 \\
 q_3 &  q_0 & -q_1 \\
-q_2 &  q_1 &  q_0
\end{array}
\right] \text{ and} \label{eq:def_B} \\
\mt C &= \left[
\begin{array}{rrrr}
-q_1 & -q_2 & -q_3 \\
 q_0 &  q_3 & -q_2 \\
-q_3 &  q_0 &  q_1 \\
 q_2 & -q_1 &  q_0
\end{array}
\right].
\end{align}
\end{subequations}

@@

The equation above is a simple and convenient factorization of matrix $\mt A$, since the entries of both $\mt B$ and $\mt C$ depend linearly on the Euler parameters. As demonstrated in Appendix~\ref{sec:auxiliary_math}, this factorization provides a relation between the angular velocity vector $\vt \omega$ and the rates of change of the Euler parameters. Such relation is
\begin{equation}
\label{eq:relation_qdot_omega}
\dot{\vt q} = \frac{1}{2} \mt B \vt \omega.
\end{equation}

As explained in the appendix, premultiplication with either $\mt B$ or $\mt C$ transforms a vector into a quaternion that lies in the hyperplane orthogonal to $\vt q$. Therefore, the equation above satisfies Eq.~\ref{eq:diff_qTq} identically and guarantees that the norm of $\vt q$ remains unchanged. Both transformations can be reversed by applying $\tr{\vt B}$ or $\tr{\vt C}$ accordingly.

@@

The relation between $\vt q$ and $\vt \omega$ (Eq.~\ref{eq:relation_qdot_omega}) obtained in the previous subsection is used here to develop a Hamiltonian formalism for the motion of rigid body. As usual, it can be described by detaching translation and rotation, represented by ${\vt r}(t)$ and ${\vt q}(t)$, respectively. The position of each particle $j$ will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$. Several authors have obtained Hamiltonian equations for the rotation of rigid bodies using quaternions as generalized coordinates.\cite{Maciejewski1985, Dichmann1996, Miller2002, Ravishankar2004, Nielsen2012} Here we try to combine convenient features of their formulations.

@@

The equations of motion are obtained from the Hamiltonian by $\dot{\vt r} = \partial \mathcal{H} / \partial \vt p$, $\dot{\vt p} = -\partial \mathcal{H} / \partial \vt r$, $\dot{\vt q} = \partial \mathcal{H} / \partial \vt \pi$, and $\dot{\vt \pi} = -\partial \mathcal{H} / \partial \vt q$.\cite{Goldstein2002} Eq.~\ref{eq:H_with_B} is convenient for differentiation with respect to $\vt \pi$, since $\mt B$ depends on $\vt q$ only. However, if we define
\[
\mt \Omega = \left[
\begin{array}{rrrr}
-\pi_1 & -\pi_2 & -\pi_3 \\
 \pi_0 & -\pi_3 &  \pi_2 \\
 \pi_3 &  \pi_0 & -\pi_1 \\
-\pi_2 &  \pi_1 &  \pi_0
\end{array}
\right],
\]
it follows that $\tr{\mt B}{\vt \pi} = \tr{\mt \Omega}{\vt q}$ and, therefore, a convenient form of $\mathcal{H}$ for differentiation with respect to $\vt q$ is
\begin{equation}
\label{eq:H_with_Omega}
\mathcal{H} = \frac{1}{2m} \tr{\vt p} \vt p + \frac{1}{8} \tr{\vt q} {\mt \Omega} {\mt I}^{-1} \tr{\mt \Omega} \vt q + U(\vt r, \vt q).
\end{equation}

For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
\[
\diff{U}{\vt r} = -\vt F = -\sum_{j=1}^{n_p} {\vt F_j},
\]
where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference (see Appendix \ref{sec:Diff_PotEng}), and thus $\vt F$ is the resultant force on the body. To obtain the derivative $\partial U/\partial \vt q$, one must take into account that the entries of $\vt q$ are not independent, due to the constraint imposed by Eq.~\ref{eq:norm_eq_1}. The correct way of performing differentiation with respect to a constrained vector is described in Ref.~\onlinecite{Schay1995}. As demonstrated in Appendix \ref{sec:Diff_PotEng}, we have
\[
\diff{U}{\vt q} = -2 \mt C \vt \tau = -2 \mt C \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
\]
where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed reference frame. Note that the product $\mt C \vt \tau$ results in a quaternion orthogonal to $\vt q$.

Finally, after carrying out the required differentiations of $\mathcal{H}$ using either Eq.~\ref{eq:H_with_B} or Eq.~\ref{eq:H_with_Omega} (see Appendix~\ref{sec:Diff_Rules}) and substituting $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B} \vt \pi = \frac{1}{2} {\mt I}^{-1} \tr{\mt \Omega} \vt q$, we obtain the following Hamiltonian system of equations of motion for a rigid body:
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = \mt F \\
&\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \\
&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega + 2 \mt C \vt \tau
\end{align}
\end{subequations}

Anticipating a feature that will be convenient in the forthcoming sections, we now consider the entries of $\vt \omega$ individually. Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$, according to Eq.~\ref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
\begin{equation}
\label{eq:omega_entry}
\omega_k = \frac{\tr{\vt \pi} \hat{\mt B}_k \vt q}{2 I_k},
\end{equation}
where $\hat{\mt B}_k$ is one of the permutation matrices presented in Appendix~\ref{sec:auxiliary_math}. In this context, the Hamiltonian of the rigid body becomes
\[
\mathcal{H} = \frac{1}{2m} \tr{\vt p} \vt p + U(\vt r, \vt q) + \frac{1}{2} \sum_{k=1}^3 I_k \omega_k^2.
\]

@@

In order to evaluate the stability of the numerical integrators, we employ a measure given by\cite{Tuckerman2010}
\begin{equation}
\label{eq:performance}
D\!E =  \frac{1}{N_\text{steps}} \sum_{k=1}^{N_\text{steps}} \left| \frac{E_{t_k} - E_0}{E_0} \right|,
\end{equation}
from which it is possible to quantify the average deviation of the supposedly conserved quantity $E$ (either $\mathcal{H}$ of Eq.~\ref{eq:H_NVE} or $H$ of Eq.~\ref{eq:H_nvt}) from its initial value $E_0$ after a certain number of time steps $N_\text{steps}$. Both the instantaneous fluctuations, given by the term inside the summation, and the long-time conservation of energy, $D\!E$, are monitored.

@@

%\begin{table}[h]
%\centering % centering table
%\begin{threeparttable}
%\caption{Partition of kinetic energy from an NVE simulation of 903 water molecules with $r_c = 9$ \AA~ and without shifting of LJ model\tnote{a}} %($5415 grados de libertad$)}
%\begin{tabular}{| l c  c  c  c  c  c |}  
%\toprule
%\hline
% & & & \multicolumn{2}{c}{Equipartition} & \multicolumn{2}{c|}{Simulated}\\
%  & & & \multicolumn{2}{c}{of energy} & \multicolumn{2}{c|}{partition}\\
%\cline{4-7}
%$\Delta t/fs $  &$T_{sim}/K$  &$E_{total}$ &$E_{trans}$ &$E_{rot}$ &$E_{trans}$ &$E_{rot}$ \\
%\hline % inserts single-line
% 0.5 & 305.04 & 1641.21  & 820.15  & 821.06  & 820.55 & 820.66  \\
%  1  & 304.88  & 1640.36  & 819.73  & 820.63  & 820.56 & 819.80  \\
%  2  & 304.04  & 1635.87  & 817.48  & 818.39  & 820.44 & 815.43 \\
%  3  & 302.78  & 1629.10  & 814.10  & 815.00  & 819.85 & 809.25 \\
%  \hline
%\end{tabular}
%\begin{tablenotes}
%\item[a] Energies are in $kcal/mol$.
%\end{tablenotes}
%\label{table:water_partition_2}
%\end{threeparttable}
%\end{adjustbox}
%\end{table}

>>>>

%\documentclass[prb,preprint,preprintnumbers,amsmath,amssymb]{revtex4-1}
%\documentclass[prb,twocolumn,preprintnumbers,amsmath,amssymb]{revtex4-1}
\documentclass[aip,jcp,reprint,amsmath,amssymb]{revtex4-1}

\usepackage[pdftex]{graphicx}
%\usepackage{epstopdf}
%\usepackage{caption}
%\usepackage{subfig}
%\usepackage[]{units}
%\usepackage[]{natbib}
\usepackage[]{threeparttable}
%\usepackage{booktabs}
%\usepackage{multirow}

\newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}           % matrix symbol
\newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}           % vector symbol
\newcommand{\tr}[1]{#1^\text{t}}                               % transposition
\newcommand{\diff}[2]{\dfrac{\partial #1}{\partial #2}} % partial derivative

\newenvironment{smallarray}[1]
{\null\,\vcenter\bgroup\scriptsize
	\renewcommand{\arraystretch}{0.7}
	\arraycolsep=.13885em
	\hbox\bgroup$\array{@{}#1@{}}}
{\endarray$\egroup\egroup\,\null}

@@

In this section, we first review certain aspects of the kinematics of rigid bodies which are relevant to the present study. Unit quaternions are employed to represent the orientations of rigid bodies. Then, we introduce the use of a particular factorization of the rotation matrix, from which a mathematical relation arises between the rate of change of a four-dimensional quaternion and a three-dimensional angular velocity vector. An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal motion.

@@

Consider a rigid body composed of $n_p$ individual particles, with each particle $j$ having mass $M_j$ and being located at a Cartesian coordinate $\vt R_j$. The center of mass of this body is located at
\[
\vt r = \frac{1}{m}\sum_{j=1}^{n_p} M_j {\vt R}_j,
\]
where $m = \sum_{j=1}^{n_p} M_j$ is the total mass of the body. The position of each particle $j$ in a space-fixed frame of reference, whose origin coincides with the body's center of mass, is ${\vt \delta}_j = \tr{ [\begin{array}{ccc} {\delta_j}_x & {\delta_j}_y & {\delta_j}_z \end{array}]} = {\vt R}_j - {\vt r}$. In such frame, the symmetric inertia tensor for rotations of the body about the origin is\cite{Goldstein2002}
\[
{\mt I}^\ast = \sum_{j=1}^N M_j \left[ \begin{smallarray}{ccc}
{\delta_j}_y^2 + {\delta_j}_z^2 & -{\delta_j}_x {\delta_j}_y      & -{\delta_j}_x {\delta_j}_z      \\
-{\delta_j}_y {\delta_j}_x      & {\delta_j}_x^2 + {\delta_j}_z^2 & -{\delta_j}_y {\delta_j}_z      \\
-{\delta_j}_z {\delta_j}_x      & -{\delta_j}_z {\delta_j}_y      & {\delta_j}_x^2 + {\delta_j}_y^2 \\
\end{smallarray} \right].
\]

For every non-collinear rigid body, there exists a particular frame of reference in which its inertia tensor is diagonal. The formulation of the equations of motion in this frame of reference is notably simplified. Conversions from the space-fixed frame to such a body-fixed frame are performed by a rotation matrix $\mt A$ that satisfies the similarity relation\citep{Goldstein2002}
\[
{\mt I} = {\mt A} {\mt I}^\ast \tr{\mt A} = \left[ \begin{smallarray}{ccc}
I_1 &   0 &   0 \\
  0 & I_2 &   0 \\
  0 &   0 & I_3 \\
\end{smallarray} \right],
\]
where $I_1$, $I_2$, and $I_3$ (the principal moments of inertia) are the eigenvalues of ${\mt I}^\ast$ and the columns of $\tr{\mt A}$ (the principal axes of rotation) are their corresponding eigenvectors. Algorithms for calculating $\mt A$ from $\mt I^\ast$ can be found in the literature.\cite{Kopp2008} Once this is done, the position of every particle $j$ in the body-fixed frame can be obtained by ${\vt d}_j = {\mt A} {\vt \delta}_j$. Since $\mt A$ represents an orthogonal transformation, a conversion in the opposite direction is done by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} This is so because $\mt A \tr{\mt A} = \mt 1_3$, where $\mt 1_k$ represents an identity matrix in $\mathbb{R}^{k \times k}$. By differentiating both sides of this relation with respect to time, we conclude that ${\mt A} \tr{\dot{\mt A}} = -\dot{\mt A} \tr{\mt A}$, which means that the result of $\mt A \tr{\dot{\mt A}}$ is skew-symmetric. The angular velocity of the body expressed in the body-fixed frame, $\vt \omega$, is related to this product by $\mt S(\vt \omega) = \mt A \tr{\dot{\mt A}}$,\cite{Haug1989} where $\mt S(\cdot)$ is the skew-symmetric operator defined in Appendix \ref{sec:auxiliary_math}  (Eq.~\ref{eq:operator_S}).

An arbitrary rotation can be described using three independent parameters. One can use, for instance, the so-called Euler angles, but mathematical singularities arise when the equations of motion are described in such terms. Instead, we consider an alternative representation, in which the orientation of a rigid body is given by the Euler parameters $q_0$, $q_1$, $q_2$, and $q_3$, which form a quaternion $\vt q = \tr {[\begin{array}{cccc} q_0 & q_1 & q_2 & q_3 \end{array}]}$. The mathematical relation between the Euler angles and the Euler parameters requires that\cite{Goldstein2002}
\begin{equation}
\label{eq:norm_eq_1}
\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_0^2 + q_1^2 + q_2^2 + q_3^2 = 1.
\end{equation}

Therefore, $\vt q$ is restricted to the surface of a unit hypersphere in $\mathbb{R}^4$, which is itself a three-dimensional domain. Appendix~\ref{sec:quat_from_A} describes how one can compute the unit quaternion related to a given rotation matrix. Reversely, the matrix $\mt A$ can be directly computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
\begin{equation}
\label{eq:A_from_q}
\mt A = \left[ \begin{smallarray}{ccc}
q_0^2 + q_1^2 - q_2^2 - q_3^2 & 2(q_1 q_2 + q_0 q_3)          & 2(q_1 q_3 - q_0 q_2) \\
2(q_1 q_2 - q_0 q_3)          & q_0^2 - q_1^2 + q_2^2 - q_3^2 & 2(q_2 q_3 + q_0 q_1) \\
2(q_1 q_3 + q_0 q_2)          & 2(q_2 q_3 - q_0 q_1)          & q_0^2 - q_1^2 - q_2^2 + q_3^2  
\end{smallarray} \right].
\end{equation}

Differentiating Eq.~\ref{eq:norm_eq_1} with respect to time yields
\begin{equation}
\label{eq:diff_qTq}
\tr{\vt q}\dot{\vt q} = 0,
\end{equation}
which implies that, for Eq.~\ref{eq:norm_eq_1} to be preserved upon an infinitesimal motion, the rate of change $\dot{\vt q}$ must lie in the hyperplane orthogonal to $\vt q$.

Let us now introduce two new matrices that also depend on the quaternion components, which are
\begin{equation}
\label{eq:def_B_and_C}
\mt B = \left[
\begin{smallarray}{rrrr}
-q_1 & -q_2 & -q_3 \\
 q_0 & -q_3 &  q_2 \\
 q_3 &  q_0 & -q_1 \\
-q_2 &  q_1 &  q_0
\end{smallarray}
\right]
\quad \text{and} \quad
\mt C = \left[
\begin{smallarray}{rrrr}
-q_1 & -q_2 & -q_3 \\
 q_0 &  q_3 & -q_2 \\
-q_3 &  q_0 &  q_1 \\
 q_2 & -q_1 &  q_0
\end{smallarray}
\right].
\end{equation}

@@

The equation above is a simple and convenient factorization of matrix $\mt A$, since the entries of both $\mt B$ and $\mt C$ depend linearly on the Euler parameters. As demonstrated in Appendix~\ref{sec:auxiliary_math}, this factorization provides a relation between the angular velocity vector $\vt \omega$ and the rates of change of the quaternion components, which is
\begin{equation}
\label{eq:relation_qdot_omega}
\dot{\vt q} = \frac{1}{2} \mt B \vt \omega.
\end{equation}

As explained in the mentioned appendix, premultiplication of a vector with either $\mt B$ or $\mt C$ transforms it into a quaternion that lies in the hyperplane orthogonal to $\vt q$. Both transformations can be reversed by applying $\tr{\vt B}$ or $\tr{\vt C}$ accordingly. Therefore, one can define an intermediary frame of reference, fixed with respect to the unit quaternion $\vt q$ that represents the orientation of the body, whose connection to the body-fixed and the space-fixed frames are summarized as follows:
\[
\boxed{\substack{\text{Body-fixed} \\ \text{frame}}}
\mathrel{\mathop{\rightleftarrows}^{\frac{1}{2}\mt B}_{\frac{1}{2}\tr{\mt B}}}
\boxed{\substack{\text{Quaternion-fixed} \\ \text{frame}}}
\mathrel{\mathop{\rightleftarrows}^{2 \tr{\mt C}}_{2 \mt C}}
\boxed{\substack{\text{Space-fixed} \\ \text{frame}}}
\]

Therefore, Eq.~\ref{eq:relation_qdot_omega} shows that $\dot{\vt q}$ is simply the angular velocity vector expressed in the quaternion-fixed frame. It also guarantees that the norm of $\vt q$ remains unchanged because Eq.~\ref{eq:diff_qTq} is identically satisfied.

@@

The relation between $\vt q$ and $\vt \omega$ (Eq.~\ref{eq:relation_qdot_omega}) is used here to develop a Hamiltonian formalism for the motion of a rigid body. As usual, it can be described by detaching translation and rotation, represented by ${\vt r}(t)$ and ${\vt q}(t)$, respectively. The position of each particle $j$ will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$. Several authors have obtained Hamiltonian equations for the rotation of rigid bodies using quaternions as generalized coordinates.\cite{Maciejewski1985, Dichmann1996, Miller2002, Ravishankar2004, Nielsen2012} Here we try to combine convenient features of their formulations.

@@

The equations of motion are obtained from the Hamiltonian by $\dot{\vt r} = \partial \mathcal{H} / \partial \vt p$, $\dot{\vt p} = -\partial \mathcal{H} / \partial \vt r$, $\dot{\vt q} = \partial \mathcal{H} / \partial \vt \pi$, and $\dot{\vt \pi} = -\partial \mathcal{H} / \partial \vt q$.\cite{Goldstein2002} Eq.~\ref{eq:H_with_B} is convenient for differentiation with respect to $\vt \pi$, since $\mt B$ depends on $\vt q$ only. However, if we define
\[
\mt \Omega = \left[
\begin{smallarray}{rrrr}
-\pi_1 & -\pi_2 & -\pi_3 \\
 \pi_0 & -\pi_3 &  \pi_2 \\
 \pi_3 &  \pi_0 & -\pi_1 \\
-\pi_2 &  \pi_1 &  \pi_0
\end{smallarray}
\right],
\]
it follows that $\tr{\mt B}{\vt \pi} = \tr{\mt \Omega}{\vt q}$ and, therefore, a convenient form of $\mathcal{H}$ for differentiation with respect to $\vt q$ is
\begin{equation}
\label{eq:H_with_Omega}
\mathcal{H} = \frac{1}{2m} \tr{\vt p} \vt p + \frac{1}{8} \tr{\vt q} {\mt \Omega} {\mt I}^{-1} \tr{\mt \Omega} \vt q + U(\vt r, \vt q).
\end{equation}

For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
\[
\diff{U}{\vt r} = -\vt F = -\sum_{j=1}^{n_p} {\vt F_j},
\]
where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference (see Appendix \ref{sec:Diff_PotEng}), and thus $\vt F$ is the resultant force on the body. To obtain the derivative $\partial U/\partial \vt q$, one must take into account that the entries of $\vt q$ are not independent, due to the constraint imposed by Eq.~\ref{eq:norm_eq_1}. The correct way of performing differentiation with respect to a constrained vector is described in Ref.~\onlinecite{Schay1995}. As demonstrated in Appendix \ref{sec:Diff_PotEng}, we have
\[
\diff{U}{\vt q} = -2 \mt C \vt \tau = -2 \mt C \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
\]
where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed reference frame. Therefore, $-\partial U/\partial \vt q$ is simply the torque represented in the quaternion-fixed frame. Finally, after carrying out the required differentiations of $\mathcal{H}$ using either Eq.~\ref{eq:H_with_B} or Eq.~\ref{eq:H_with_Omega} (see Appendix~\ref{sec:Diff_Rules}) and substituting $\tr{\mt B} \vt \pi = \tr{\mt \Omega} \vt q = 2 {\mt I} \vt \omega$, we obtain the following Hamiltonian system of equations of motion for a rigid body:
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = \mt F \\
&\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \\
&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega + 2 \mt C \vt \tau \label{eq:euler_eq_quat}
\end{align}
\end{subequations}

In Appendix \ref{eq:derivation_euler}, it is demonstrated that Eq.~\ref{eq:euler_eq_quat} is a quaternion-fixed frame representation of the well-known Euler equation.\cite{Goldstein2002} Anticipating a feature that will be convenient in the forthcoming sections, we now consider the entries of $\vt \omega$ individually. Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$, according to Eq.~\ref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
\begin{equation}
\label{eq:omega_entry}
\omega_k = \frac{\tr{\vt \pi} \hat{\mt B}_k \vt q}{2 I_k},
\end{equation}
where $\hat{\mt B}_k$ is one of the permutation matrices presented in Appendix~\ref{sec:auxiliary_math}. In this context, the Hamiltonian of the rigid body becomes
\[
\mathcal{H} = \frac{1}{2m} \tr{\vt p} \vt p + U(\vt r, \vt q) + \frac{1}{2} \sum_{k=1}^3 I_k \omega_k^2.
\]

@@

In order to evaluate the stability of the numerical integrators, we employ a measure given by\cite{Tuckerman2010}
\begin{equation}
\label{eq:performance}
D\!E =  \frac{1}{\mathcal N} \sum_{t=1}^{\mathcal N} \left| \frac{E_t - E_0}{E_0} \right|,
\end{equation}
from which it is possible to quantify the average deviation of the supposedly conserved quantity $E$ (either $\mathcal{H}$ of Eq.~\ref{eq:H_NVE} or $H$ of Eq.~\ref{eq:H_nvt}) from its initial value $E_0$ after a certain number of time steps $\mathcal N$. Both the instantaneous fluctuations, given by the term inside the summation, and the long-time conservation of energy, $D\!E$, are monitored.

@@

\begin{table}
	\begin{threeparttable}
		\caption{Kinetic energy partition in NVE simulations of 903 water molecules with $r_c$ = 9~{\r A} and a shifted-force Lennard-Jones potential.\tnote{a}} %($5415 grados de libertad$)}
		\label{table:water_partition_LJ-SF}
		\begin{ruledtabular}
			\begin{tabular}{ccccccc}
				& & & \multicolumn{2}{c}{Equipartitioned} & \multicolumn{2}{c}{Simulated}\\
%				& & & \multicolumn{2}{c}{of energy} & \multicolumn{2}{c}{partition}\\
				\cline{4-7}
				$\Delta t$/fs & $T_\text{sim}$/K & $\langle E_\text{kin}\rangle$ & $\langle E_\text{trans}\rangle$ & $\langle E_\text{rot}\rangle$ & $\langle E_\text{trans}\rangle$ & $\langle E_\text{rot}\rangle$ \\
				\hline
				0.5 & 305.20  & 1642.11  & 820.60  & 821.06  & 820.60 & 821.51  \\
				1.0 & 304.92  & 1640.57  & 819.83  & 820.74  & 820.63 & 819.94  \\
				2.0 & 304.20  & 1636.74  & 817.92  & 818.82  & 820.56 & 816.18  \\
				3.0 & 302.97  & 1630.07  & 814.59  & 815.48  & 820.47 & 809.60
			\end{tabular}
		\end{ruledtabular}
		\begin{tablenotes}
			\item[a] Energies are in kcal/mol.
		\end{tablenotes}
	\end{threeparttable}
\end{table}

\begin{table}
	\begin{threeparttable}
		\caption{Kinetic energy partition in NVE simulations of 903 water molecules with $r_c$ = 13~{\r A} and a shifted-force Lennard-Jones potential.\tnote{a}} %($5415 grados de libertad$)}
		\label{table:water_partition_13A_LJ-SF}
		\begin{ruledtabular}
			\begin{tabular}{ccccccc}
				& & & \multicolumn{2}{c}{Equipartitioned} & \multicolumn{2}{c}{Simulated}\\
%				& & & \multicolumn{2}{c}{of energy} & \multicolumn{2}{c}{partition}\\
				\cline{4-7}
				$\Delta t$/fs & $T_\text{sim}$/K & $\langle E_\text{kin}\rangle$ & $\langle E_\text{trans}\rangle$ & $\langle E_\text{rot}\rangle$ & $\langle E_\text{trans}\rangle$ & $\langle E_\text{rot}\rangle$ \\
				\hline
				0.5 & 307.17 & 1652.71  & 825.90  & 826.81  & 825.96 & 826.75  \\
				1.0 & 306.98 & 1651.68  & 825.38  & 826.30  & 826.09 & 825.59  \\
				2.0 & 306.04 & 1646.60  & 822.85  & 823.75  & 825.57 & 821.03  \\
				3.0 & 304.79 & 1639.89  & 819.49  & 820.40  & 825.63 & 814.26
			\end{tabular}
		\end{ruledtabular}
		\begin{tablenotes}
			\item[a] Energies are in kcal/mol.
		\end{tablenotes}
	\end{threeparttable}
\end{table}

\subsection{Sampling Validation}
\label{sec:samplingvalidation}

In this subsection we present the results from a simple method that serves for checking the consistency between the sampled energy distributions and the theoretical ones. The procedure, introduced and implemented by Shirts ~\cite{Shirts2013}, relies on a quantitative statistical analysis of fitting and it is possible to perform linear and nonlinear fitting or a maximum likelihood approach too. This last approach does not involve any binning of the energies, hence no discretization error is present. We report the results corresponding to that method, and since it is instructive to have a graphical insight, we also show some figures of the linear fitting.
